---
title: "In_class_Ex2"
editor: visual
---

## Overview

## Getting Started

Setting Up working Environment

```{r}
pacman::p_load(sf, tidyverse, tmap, spdep, funModeling)
```

## Importing data

-   geo_export

-   nga_admbnda_adm2_osgof_20190417

Importing water point geospatial data

```{r}
wp <- st_read(dsn = "geodata",
              layer = "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

```{r}
write_rds(wp, "geodata/wp_nga.rds")
```

Importing Nigeria LGA boundary data

```{r}
nga <- st_read(dsn = "geodata",
               layer = "nga_admbnda_adm2_osgof_20190417",
               crs = 4326)
```

## Data Wranging

Recording NA values into string

```{r}
wp_nga <- read_rds("geodata/wp_nga.rds") %>%
  mutate(status_cle = replace_na(status_cle, "Unknown"))
```

EDA

```{r}
freq(data=wp_nga, 
     input = 'status_cle')
```

## Extracting Water Point Data

Extracting functional water point

```{r}
wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional", 
             "Functional but not in use",
             "Functional but needs repair"))
```

```{r}
freq(data=wpt_functional, 
     input = 'status_cle')
```

Extracting non-functional water point

```{r}
wpt_nonfunctional <- wp_nga %>%
  filter(status_cle %in%
           c("Abandoned/Decommissioned", 
             "Abandoned",
             "Non-Functional",
             "Non functional due to dry season",
             "Non-Functional due to dry season"))
```

```{r}
freq(data=wpt_nonfunctional, 
     input = 'status_cle')
```

Extracting water point with Unknown class

```{r}
wpt_unknown <- wp_nga %>%
  filter(status_cle == "Unknown")
```

## Performing Point-in-Polygon Count

```{r}
nga_wp <- nga %>% 
  mutate(`total wpt` = lengths(
    st_intersects(nga, wp_nga))) %>%
  mutate(`wpt functional` = lengths(
    st_intersects(nga, wpt_functional))) %>%
  mutate(`wpt non-functional` = lengths(
    st_intersects(nga, wpt_nonfunctional))) %>%
  mutate(`wpt unknown` = lengths(
    st_intersects(nga, wpt_unknown)))
```

## Saving the Analytical Data Table

```{r}
nga_wp <- nga_wp %>%
  mutate(pct_functional = `wpt functional`/`total wpt`) %>%
  mutate(`pct_non-functional` = `wpt non-functional`/`total wpt`) %>%
  select(3:4, 9:10, 18:23)
```

Save the sf data table into rds format

```{r}
write_rds(nga_wp, "geodata/nga_wp.rds")
```

Visualising the spatial distribution of water points

```{r}

nga_wp <- read_rds("geodata/nga_wp.rds")
total <- qtm(nga_wp, "total wpt")+
  tm_layout(legend.height=0.25,
            legend.width=0.35)
wp_functional <- qtm(nga_wp, "wpt functional")+
  tm_layout(legend.height=0.25,
            legend.width=0.35)
wp_nonfunctional <- qtm(nga_wp, "wpt non-functional")+
  tm_layout(legend.height=0.25,
            legend.width=0.35)
unknown <- qtm(nga_wp, "wpt unknown")+
  tm_layout(legend.height=0.25,
            legend.width=0.35)

tmap_arrange(total, wp_functional, wp_nonfunctional, unknown, asp=1, ncol=2)

```
